---
title: "TransitCare Fleet Maintenance Analysis"
subtitle: "Exploratory Data Analysis for Predictive Maintenance"
format:
  html:
    theme: [default, _brand.yml]
    toc: true
    toc-depth: 3
    code-fold: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
library(dplyr)
library(ggplot2)
library(lubridate)
library(readr)
library(tidyr)
library(gt)

theme_set(theme_minimal(base_size = 12))

if (!file.exists("data/synthetic-vehicles.csv")) {
  source("data/generate_data.R")
}

vehicles <- read_csv("data/synthetic-vehicles.csv", show_col_types = FALSE)
maintenance <- read_csv("data/synthetic-maintenance.csv", show_col_types = FALSE)
vehicle_health <- read_csv("data/synthetic-vehicle-health.csv", show_col_types = FALSE)
sensor_readings <- read_csv("data/synthetic-sensor-readings.csv", show_col_types = FALSE)
```

## Executive Summary

This analysis examines TransitCare's fleet of `r nrow(vehicles)` vehicles across `r n_distinct(vehicles$depot)` depots to identify patterns in maintenance events and predict potential failures. Key findings support the development of a predictive maintenance strategy to reduce downtime and optimize fleet operations.

## Fleet Overview

### Fleet Composition

```{r}
#| label: fleet-composition
vehicles |>
  count(vehicle_type, depot) |>
  pivot_wider(names_from = depot, values_from = n, values_fill = 0) |>
  gt() |>
  tab_header(title = "Fleet Distribution by Type and Depot") |>
  cols_label(vehicle_type = "Vehicle Type")
```

```{r}
#| label: fleet-age
#| fig-height: 4
vehicles |>
  mutate(age = 2024 - year_manufactured) |>
  ggplot(aes(x = age, fill = vehicle_type)) +
  geom_histogram(binwidth = 1, position = "stack", color = "white") +
  labs(
    title = "Fleet Age Distribution",
    x = "Vehicle Age (Years)",
    y = "Number of Vehicles",
    fill = "Vehicle Type"
  ) +
  scale_fill_brewer(palette = "Set2")
```

### Vehicle Health Distribution

```{r}
#| label: health-scores
#| fig-height: 4
vehicle_health |>
  ggplot(aes(x = health_score, fill = failure_risk)) +
  geom_histogram(binwidth = 5, color = "white") +
  labs(
    title = "Vehicle Health Score Distribution",
    x = "Health Score",
    y = "Number of Vehicles",
    fill = "Failure Risk"
  ) +
  scale_fill_manual(values = c(
    "Low" = "#2ECC71",
    "Medium" = "#F39C12",
    "High" = "#E74C3C",
    "Critical" = "#8E44AD"
  ))
```

## Maintenance Analysis

### Maintenance Events by Component

```{r}
#| label: maintenance-by-component
#| fig-height: 5
maintenance |>
  count(component, severity) |>
  ggplot(aes(x = reorder(component, n), y = n, fill = severity)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Maintenance Events by Component",
    x = NULL,
    y = "Number of Events",
    fill = "Severity"
  ) +
  scale_fill_manual(values = c(
    "Minor" = "#3498DB",
    "Moderate" = "#F39C12",
    "Major" = "#E74C3C",
    "Critical" = "#8E44AD"
  ))
```

### Maintenance Cost Analysis

```{r}
#| label: cost-summary
maintenance |>
  mutate(total_cost = parts_cost + labor_cost) |>
  group_by(component) |>
  summarise(
    events = n(),
    avg_cost = mean(total_cost),
    total_cost = sum(total_cost),
    avg_downtime = mean(downtime_hours),
    .groups = "drop"
  ) |>
  arrange(desc(total_cost)) |>
  gt() |>
  tab_header(title = "Maintenance Cost Summary by Component") |>
  fmt_currency(columns = c(avg_cost, total_cost), currency = "GBP") |>
  fmt_number(columns = avg_downtime, decimals = 1) |>
  cols_label(
    component = "Component",
    events = "Events",
    avg_cost = "Avg Cost",
    total_cost = "Total Cost",
    avg_downtime = "Avg Downtime (hrs)"
  )
```

### Monthly Maintenance Trends

```{r}
#| label: monthly-trends
#| fig-height: 4
maintenance |>
  mutate(month = floor_date(date, "month")) |>
  count(month, scheduled) |>
  ggplot(aes(x = month, y = n, color = scheduled)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Monthly Maintenance Events",
    x = NULL,
    y = "Number of Events",
    color = "Scheduled"
  ) +
  scale_color_manual(values = c("TRUE" = "#2ECC71", "FALSE" = "#E74C3C"),
                     labels = c("TRUE" = "Scheduled", "FALSE" = "Unscheduled"))
```

## Predictive Indicators

### Health Score vs Failure Probability

```{r}
#| label: health-vs-failure
#| fig-height: 4
vehicle_health |>
  ggplot(aes(x = health_score, y = estimated_failure_probability, color = vehicle_type)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "loess", se = FALSE, color = "black", linetype = "dashed") +
  labs(
    title = "Health Score vs Estimated Failure Probability",
    x = "Health Score",
    y = "Failure Probability",
    color = "Vehicle Type"
  ) +
  scale_color_brewer(palette = "Set2")
```

### Key Risk Factors

```{r}
#| label: risk-factors
vehicle_health |>
  filter(failure_risk %in% c("High", "Critical")) |>
  select(vehicle_id, vehicle_type, depot, age_years, health_score,
         critical_issues_ytd, days_since_last_service, estimated_failure_probability) |>
  arrange(desc(estimated_failure_probability)) |>
  head(10) |>
  gt() |>
  tab_header(title = "Top 10 High-Risk Vehicles") |>
  fmt_percent(columns = estimated_failure_probability, decimals = 1) |>
  cols_label(
    vehicle_id = "Vehicle",
    vehicle_type = "Type",
    depot = "Depot",
    age_years = "Age",
    health_score = "Health",
    critical_issues_ytd = "Critical Issues",
    days_since_last_service = "Days Since Service",
    estimated_failure_probability = "Failure Prob"
  )
```

## Sensor Data Insights

```{r}
#| label: sensor-trends
#| fig-height: 5
sensor_readings |>
  select(vehicle_id, reading_date, engine_temp, oil_pressure, battery_voltage, brake_wear_pct) |>
  pivot_longer(cols = c(engine_temp, oil_pressure, battery_voltage, brake_wear_pct),
               names_to = "metric", values_to = "value") |>
  ggplot(aes(x = reading_date, y = value)) +
  geom_line(aes(group = vehicle_id), alpha = 0.3) +
  stat_summary(fun = mean, geom = "line", color = "#E74C3C", linewidth = 1) +
  facet_wrap(~metric, scales = "free_y", ncol = 2) +
  labs(
    title = "Sensor Reading Trends (30-Day Window)",
    subtitle = "Individual vehicles (grey) with fleet average (red)",
    x = NULL,
    y = "Value"
  )
```

## Conclusions and Recommendations

Based on this analysis:

1. **Component Focus**: Brakes and Electrical systems account for the highest maintenance frequency and should be prioritized for predictive monitoring

2. **Fleet Renewal**: Vehicles older than 7 years show significantly higher maintenance costs and failure rates

3. **Unscheduled Maintenance**: Approximately 40% of maintenance events are unscheduled, indicating opportunity for improved predictive capabilities

4. **Depot Variations**: Maintenance patterns vary by depot, suggesting potential process improvements through cross-depot learning

---

*This analysis uses synthetic data generated for demonstration purposes. All company names, vehicle identifiers, and metrics are fictional.*
